---
title: "Introduction to R and RStudio Notes"
output: statsr:::statswithr_lab
---

## RStudio Overview
R is the name of the programming language itself and RStudio is a convenient interface.

Your RStudio window has four panels:
<div>
  1. **R Markdown Panel**  (this document) in the upper left panel. Is the "master file" 
of notes, commands, and outputs. Different texts can be formatted (titles, bold,
italicized, etc) and be generated into a html file. 
<div>
  2. **Console Panel** in the bottom left has the same text at the top of the console telling 
you the version of R that you're running. Below that information is the *prompt* or 
request for command.
<div>
  3. **Workspace Panel** in the upper right has a history of commands previously entered 
and datasets that are being used. 
<div>
4. **Lower right panel** shows the plots you generate and also  where you can browse 
your files, access help, manage packages, etc.


## R Packages

R is an open-source programming language, meaning that users can contribute
packages that make our lives easier, and we can use them for free. For this lab,
and many others in the future, we will use the following R packages:

- `statsr`: for data files and functions used in this course
- `dplyr`: for data wrangling
- `ggplot2`: for data visualization

First install these packages: <div>
- install.packages("dplyr") 
- install.packages("ggplot2")
- install.packages("shiny")
- install_github("StatsWithR/statsr")

Next, you need to load the packages in your working environment with the `library` 
function. Note that you only need to **install** packages once, but
you need to **load** them each time you relaunch RStudio.

```{r load-packages, message = FALSE}
library(dplyr)
library(ggplot2)
library(statsr)
```

Three ways to run the commands: <div>
1. Highlight the codes to run -> Run -> Run current Chunk 
2. Type the code into the console
3. Run to run the entire rmd file

## Simple Data Commands

*About the Data:* The data set refers to Dr. John Arbuthnot, an 18<sup>th</sup> century 
physician, writer, and mathematician. He was interested in the ratio of newborn
boys to newborn girls, so he gathered the baptism records for children born in
London for every year from 1629 to 1710. 

**Load Data Sets **
```{r load-abrbuthnot-data}
data(arbuthnot)
```
Data is successfully loaded when it appears in the Workspace Panel as 'arbuthnot'
with 82 observations and 3 variables. 

**View Data**

```{r view-data}
arbuthnot
```

**View Dimensions of a DataFrame**

```{r dim-data}
dim(arbuthnot)
```

Output indicates that there are 82 rows and 3 columns  

**View Names of Columns of Dataset**

```{r names-data}
names(arbuthnot)
```

**View Data in a Column**

```{r year-data}
arbuthnot$year
```

This command will only show the years baptized. The dollar sign basically says 
"go to the data frame that comes before me, and find the variable that comes after me".

**Create Scatter Plot**

```{r plot-girls-vs-year}
ggplot(data = arbuthnot, aes(x = year, y = girls)) +
  geom_point()
```

Use  `ggplot()` function to build plots. You should see the plot appear under the *Plots* tab 
of the lower right panel of RStudio. <div>
- The first argument is always the dataset. 
- Then provide the variables from the dataset to be assigned to `aes`thetic 
elements of the plot, e.g. the x and the y axes. 
- Finally, we use another layer, separated by a `+` to specify the `geom`etric 
object for the plot. Since we want to scatterplot, we use `geom_point`.

**Find Synatx for a Function**
To find more syntax for the `ggplot` function just type in a question mark 
followed by the name of the function that you're interested in.  

```{r plot-help, tidy = FALSE}
?ggplot
```

Notice that the help file replaces the plot in the lower right panel. You can 
toggle between plots and help files using the tabs at the top of that panel. 

More extensive help for plotting with the `ggplot2` package can be found at 
http://docs.ggplot2.org/current/. 


### R as a big calculator

**Simple Addition**

```{r calc-total-bapt-numbers}
5218 + 4683
```

**Add Vectors**

```{r calc-total-bapt-vars}
arbuthnot$boys + arbuthnot$girls
```

### Adding a new variable to the data frame

```{r calc-total-bapt-vars-save}
arbuthnot <- arbuthnot %>%
  mutate(total = boys + girls)
```

The `%>%` operator is called the **piping** operator. It takes the output of the 
current line and pipes it into the following line of code.

  It can be read as *"Take the `arbuthnot` dataset and **pipe** it into the `mutate` 
  function. Mutate a new variable called `total` that is the sum of the variables
  called `boys` and `girls`. Then overwrite the old `arbuthnot` dataset with the new 
  arbuthnot dataset containing the new variable (total)."*
    
  This is essentially equivalent to going through each row and adding up the boys 
  and girls counts for that year and recording that value in a new column called
  total.

`<-` performs an *assignment*
    takes the output of one line of code and saving it into an object in your workspace. In 
    this case, you already have an object called `arbuthnot`, so this command updates
    that data set with the new mutated column.

** Create a Line Graph**
```{r plot-total-vs-year-line}
ggplot(data = arbuthnot, aes(x = year, y = total)) +
  geom_line()
```

Note that using `geom_line()` instead of `geom_point()` results in a line plot instead
of a scatter plot. You want both? Just layer them on:

```{r plot-total-vs-year-line-and-point}
ggplot(data = arbuthnot, aes(x = year, y = total)) +
  geom_line() +
  geom_point()
```

**Graph Using Arithmetics**
```{r plot-proportion-of-boys-over-time}
ggplot(data = arbuthnot, aes(x = year, y = boys/total)) +
  geom_line() +
  geom_point()
```

**Graph using Comparisons**
```{r boys-more-than-girls}
arbuthnot <- arbuthnot %>%
  mutate(more_boys = boys > girls)

ggplot(data = arbuthnot, aes(x = year, y = more_boys)) +
  geom_point()
```

R can make comparisons like greater than, `>`, less than,`<`, and equality, `==`. 
For example, we can ask if boys outnumber girls in each year with the expression
This command add a new variable to the `arbuthnot` data frame containing the values
of either `TRUE` if that year had more boys than girls, or `FALSE` if that year 
did not (the answer may surprise you). Here,we've asked R to create *logical* data.


## Resources for learning R and working in RStudio
- The following cheathseets may come in handy throughout the course. Note that some 
of the code on these cheatsheets may be too advanced for this course, however 
majority of it will become useful as you progress through the course material.
    - [Data wrangling cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
    - [Data visualization cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)
    - [R Markdown](http://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)

- While you will get plenty of exercise working with these packages in the labs of 
this course, if you would like further opportunities to practice we recommend 
checking out the relevant courses at [DataCamp](https://www.datacamp.com/courses).

<div id="license">
This is a derivative of an [OpenIntro](https://www.openintro.org/stat/labs.php) lab, and is released under a [Attribution-NonCommercial-ShareAlike 3.0 United States](https://creativecommons.org/licenses/by-nc-sa/3.0/us/) license.
</div>