---
title: "Normal Distribution and Binomial Distribution"
output: statsr:::statswithr_lab
---
___
### Normal Distribution
___
**Question 1**
Suppose weights of checked baggage of airline passengers follow a nearly normal distribution with mean 45 pounds and standard deviation 3.2 pounts. Most arlines charge a fee for baggage that weigh in excess of 50 pounds. What percent of airline passengers are expected to incur this fee? 

*Method 1: R*
```{r percentile-pnorm}
pnorm(50, mean=45, sd = 3.2)
```

**_NOTE:_** pnorm provides the cumulative density of the normal distribution at a specific quantile. Area under the curve aka the "cumulative density". <div>

\
*Method 2: Hand Calculation*<div>
1. Calculate z score
    z = (observed value - mean)/standard deviation
      = (50-45)/3.2 
      = 1.56
2. Find the z value in the z-chart 
    P(z) ~ 0.9406
3. Calculating in **excess** of 50
    P( z>1.56 ) = 1-0.9406 = 0.0594

*Method 3: Applet*

___
**Question 2**

The average daily high temperature in June in LA is 77 with a standard deviation of 5. Suppose that the temperatures in June closely follow a normal distribution. How cold are the coldest 20% of the days during June in LA? 

*Method 1: R*
```{r percentile-qnorm}
qnorm(0.20,mean=77,sd=5)
```

qnorm provides the quantile of the normal distribution at a specified cumulative density. Aka finding the number on the bottom. 

*Method 2: Hand Calculation*

1. Find the z value in the z-chart 
    0.20 percentile corresponds to a z-value of -0.84
2. Calculate observed value using z-score
        z = (observed value - mean)/standard deviation
    -0.84 = (x-77)/5
        x = 72.8
        
*Method 3: Applet*

___
### Binomial Distribution
___

According to a 2013 Gallup poll, worldwide only 13% of employees are engaged at work. AMong a random sample of 10 employees, what is the probability that 8 of them are engaged at work? 
  
- n=10         
- k=8
- p=0.13     
- 1-p=0.87    

```{R binomial-distribution}
dbinom(8, size=10, p=0.13)
```







___
**Load Data**

```{r load-data}
library(dplyr)
library(ggplot2)
library(statsr)
```